Simulating High-Dimensional Multivariate Data

### 2020-12-19 010-background.Rmd
4th draft. Address JP feedback

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (BP) and "scaleability" properties (SP). 
The BPs are adapted from an existing criteria due to @Nik13a.
Our simulation strategy should allow:

@@

To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

\[
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
\]

A Gaussian copula has marginal CDFs as the normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

>>>>

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP). 
The BPs are adapted from an existing criteria due to @Nik13a.
A suitable simulation strategy should possess the following properties:

@@

<!-- 
The next section describes the motivating data set we use to provide example applications.
your comment -->

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
Importantly, due to inherent biological processes, gene expression data exhibit correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scalable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

\begin{equation}
(\#eq:copula)
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\end{equation}

$x_i \in \mathbb R, i=1,\ldots,d.$

A Gaussian copula has marginal CDFs that are all normal, $F_i = \Phi, \forall \, i$. 
This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

### 2020-12-19 020-simulation-algorithm.Rmd
4th draft. Address JP feedback

The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
Specifically, NORTA algorithm proceedings as follows:

@@

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing fast scaleability using readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely, see @Chen2001; @Xia17; @MB13 and the publicly available `nortaRA` R package.

To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint.
Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed below in Section \@ref(rand-vec-gen)).

@@

## Random vector generation via `bigsimr::rvec` {#rand-vec-gen}

We now describe `bigsimr::rvec`, our algorithm to generate random vectors.
It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not, compute the nearest correlation matrix ${\bf \widetilde{R}_{Pearson}}$ to ${\bf R_{Pearson}}$ in the Frobenius norm.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
We hope that this routine will be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm from the preceding section).

@@

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. 
They found at what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

>>>>

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
Specifically, NORTA algorithm proceedings as follows:

@@

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic.
In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$.
Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$.
For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies.
For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].

To maintain scalability (SP1), our solution is to essentially avoid this complication in Pearson matching.
Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures, such as $\rho_S$ and $\tau$, from a modeling standpoint.
Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed below in Section \@ref(rand-vec-gen)).

@@

## `bigsimr::rvec` for random vector generation  {#rand-vec-gen}

We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert ${\bf R_{Kendall}}$ or ${\bf R_{Spearman}}$ to the corresponding MVN correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not, compute the nearest correlation matrix ${\bf \widetilde{R}_{Pearson}}$ to ${\bf R_{Pearson}}$ in the Frobenius norm.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
This routine as many applications beyond our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, Steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication.

@@

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. 
They found for what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

### 2020-12-19 030-bigsimr-package.Rmd
4th draft. Address JP feedback

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
This implementation strategy supports all `R` `base` probability distributions.
It allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.

@@

*Checking the theoretical correlation bounds*.
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
To ensure that the simulation is not configured to impossible settings, the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).

@@

It is important to follow `R`'s naming convention for probability distributions.
Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, one must to supply a univariate random number generator as well.

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Though `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
A single run of `bigsimr:rvec` using `rslurm` can be coded as:

>>>>

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ are specified the `bigsimr::rvec` algorithm.
This implementation strategy supports all `R` `base` probability distributions.
It allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

To avoid manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.

@@

*Checking the theoretical correlation bounds*.
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
`bigsimr::cor_bounds` estimates the theoretic lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.

@@

Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, one must to supply a univariate random number generator as well.

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
The R package `rslurm` enables parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
A single run of `bigsimr:rvec` using `rslurm` can be coded as:

### 2020-12-19 040-monte-carlo-evaluation.Rmd
4th draft. Address JP feedback

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example and, so, proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). 
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. 
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

@@

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
Taken together, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

>>>>

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \@ref(examples). 
We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations.
For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. 
The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

@@

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

### 2020-12-19 050-examples.Rmd
4th draft. Address JP feedback

Next, we estimate the marginal parameters. 
We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
To conclude the example applications, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
In particular, we wish to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory, which unfortunately, was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (.e.g., @Won2013g and @VanWieringen2016).

>>>>

Next, we estimate the marginal parameters. 
We use the method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below support this estimation for later use in `bigsimr::rvec`.

@@

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
To conclude the example applications, we demonstrate how `bigsimr` can be used to evaluate estimation efficiency.
In particular, we wish to assess the error in our correlation estimation above.
We used a conventional method, based on classical statistical theory, which unfortunately, was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (.e.g., @Won2013g and @VanWieringen2016).

### 2020-12-19 060-conclusion-discussion.Rmd
4th draft. Address JP feedback

We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
The major contributions of this work are high-dimensional scaleability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.

>>>>

We have introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package named [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.

### 2020-12-19 070-supplementary-materials.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1
4th draft. Remove naive References heading
4th draft. Address JP feedback

The authors gratefully acknowledge the helpful discussions with University of Arizona's Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
We also gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.

@@

Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.

>>>>

The authors gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.

@@

Research reported in this publication was supported by grants from NV INBRE and the National Institutes of General Medical Sciences (NIH 5U54GM104944, GM103440).
